---
title: "Lisa Vegetable Garden"
description: |
  Analysis of Lisa Vegetable Garden data from TidyTuesday
author: Ian Palmer
date: September 18, 2024
format: html
execute: 
  warning: false
  message: false
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

This project showcases my first time using the introductory tidyverse data visualization skills that I learned in my data science course. In this section I looked at the Lisa Vegetable Garden data set from tidytuesday. I aimed to explore the data set and create a visualization using the data.

The dataset used in this project, **Lisa's Vegetable Garden**, is sourced from [Tidy Tuesday's May 28, 2024 release](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-05-28/readme.md).

The Lisa's Vegetable Garden data set contains detailed records of Lisa Lendway's vegetable garden from the summers of 2020 and 2021. The data includes planting, harvesting, and spending details, offering insights into gardening practices and changes between the two years. Originally used in her Introduction to Data Science course at Macalester College, the data set provides a hands-on way to explore data science concepts.

Tidy Tuesday is a weekly R project encouraging data wrangling, visualization, and sharing insights with the #TidyTuesday community.

Here is the dataset that I used:

```{r, message=FALSE, warning=FALSE}

tuesdata <- tidytuesdayR::tt_load(2024, week = 22)

harvest_2020 <- tuesdata$harvest_2020
harvest_2021 <- tuesdata$harvest_2021
planting_2020 <- tuesdata$planting_2020
planting_2021 <- tuesdata$planting_2021
spending_2020 <- tuesdata$spending_2020
spending_2021 <- tuesdata$spending_2021
```

Here is a quick glimpse of the data contained in the harvest_2020 data set.

```{r}
head(harvest_2020,5)

```

This code chuck creates a new data set called harvest_2020_avg which I use in my visualization later. I find the average harvest weight across all months in the 2020 harvest season.

```{r}
# Create a new dataset with average harvest weight per month
harvest_2020_avg <- harvest_2020 |>
  # Add a new column 'month', extracting the year and month from the 'date' column
  mutate(month = as.Date(format(as.Date(date), "%Y-%m-01"))) |> 
  # Group the data by the 'month' column
  group_by(month) |>
  # Calculate the average weight for each month, ignoring missing values (NA)
  summarize(avg_weight = mean(weight, na.rm = TRUE))

# Display the resulting dataset
harvest_2020_avg

```

Below you can see my code and data visualization of Lisa's Vegetable Garden. The graph represents the average weight of vegetables harvested during each month of the 2020 harvest.

```{r, fig.alt= "This line graph shows the average weight of vegetables harvested per month during the 2020 season. The x-axis represents the months (June to October), and the y-axis represents the average weight of the harvest in grams. The trend line starts low in June, rises steadily through July and August, peaks in September, and then slightly declines in October. Each data point is marked with a dot connected by straight lines."}


# Create a line plot of average vegetable weights harvested per month
ggplot(harvest_2020_avg, aes(x = month, y = avg_weight)) +  # Set up ggplot with dataset and aesthetic mappings
  geom_line() +  # Add a line connecting the average weights for each month
  geom_point() +  # Add points at each month's average weight for emphasis
  labs(
    title = "Average Weight of Vegetables Harvested per Month in 2020 Season",  # Set the plot title
    x = "Month",  # Label for the x-axis (Month)
    y = "Average Weight (grams)"  # Label for the y-axis (Average Weight in grams)
  )

  
```

As we can see, the graph shows the average weight of vegetables harvested per month during the 2020 season. Harvest weights increased steadily from June, peaked in September, and then slightly declined in October, highlighting a clear seasonal trend in vegetable yields.
