---
title: "Mini Project 3 Final"
author: "Ian Palmer"
date: 11/08/24
execute: 
  warning: false
  message: false
---
Analysis Plan:

To test whether home-court advantage has a measurable effect on win percentages, we’ll calculate the observed difference in median win percentage between home and away games across all NCAA teams in our dataset. To determine if this observed difference could happen by chance, we’ll use a permutation test. Specifically, we’ll shuffle the "home" and "away" labels for each team multiple times and calculate the median difference in win percentages after each shuffle. By comparing our observed difference to this distribution of permuted differences, we can assess whether any observed effect is statistically significant.

The result will show whether the median win percentage difference between home and away games is greater than what we’d expect by random chance alone. A low p-value would suggest that the difference is meaningful and likely due to a true home-court advantage.

Data Source:

The data used in this analysis comes from a Kaggle dataset by Nishaan Amin, titled March Madness Data, which includes statistics on NCAA men's basketball tournament teams from 2008 to 2024 (excluding 2020, as the tournament was canceled due to COVID-19). Specifically, we used two files: Barttorvik Home.csv and Barttorvik Away.csv, which provide team statistics for home and away games. These files were sourced from [Barttorvik’s college basketball data](https://www.barttorvik.com/#). The full dataset is available on Kaggle: [Home file](https://www.kaggle.com/datasets/nishaanamin/march-madness-data/data?select=Barttorvik+Home.csv) and [Away file](https://www.kaggle.com/datasets/nishaanamin/march-madness-data/data?select=Barttorvik+Away.csv).


```{r}
library(tidyverse)
# Original Data
home_data <- read.csv("Barttorvik Home.csv")
away_data <- read.csv("Barttorvik Away.csv")

```


```{r}
# Select Relevant Columns and Add Location Column
home_data <- home_data |>
  select(TEAM, WIN.) |>
  rename(win_pct = WIN.) |>
  mutate(location = "home")

away_data <- away_data |>
  select(TEAM, WIN.) |>
  rename(win_pct = WIN.) |>
  mutate(location = "away")

#Group by TEAM and Location, Calculate Median Win Percentage
home_median <- home_data |>
  group_by(TEAM, location) |>
  summarise(median_win_pct = median(win_pct, na.rm = TRUE))

away_median <- away_data |>
  group_by(TEAM, location) |>
  summarise(median_win_pct = median(win_pct, na.rm = TRUE))

# Combine Home and Away Results
combined_median_data <- bind_rows(home_median, away_median)
```

```{r}
#Calculate the Observed Difference in Median Win Percentage
observed_diff <- combined_median_data |>
  group_by(location) |>
  summarise(median_win_pct = median(median_win_pct, na.rm = TRUE)) |>
  summarise(diff = diff(median_win_pct)) |>
  pull(diff)

cat("Observed Difference in Median Win Percentage (Home - Away):", observed_diff, "\n")
```

```{r}
# Define Permutation Function (Shuffling location within each team)
calculate_permutation <- function(data) {
  data |>
    group_by(TEAM) |>
    mutate(location = sample(location, replace = FALSE)) |>
    group_by(location) |>
    summarise(median_win_pct = median(median_win_pct, na.rm = TRUE)) |>
    summarise(diff = diff(median_win_pct)) |>
    pull(diff)
}
#Permutation Test
num_permutations <- 10000
perm_results <- map_dbl(1:num_permutations, ~ calculate_permutation(combined_median_data))


p_value <- mean(abs(perm_results) >= abs(observed_diff))
cat("Two-Sided P-value:", p_value, "\n")
```

```{r}
# Histogram of Permuted Differences with the Observed Difference
ggplot(data.frame(perm_results), aes(x = perm_results)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  geom_vline(xintercept = observed_diff, color = "red", linetype = "dashed", linewidth = 1.2) +
  labs(
    title = "Permutation Test: Distribution of Permuted Differences",
    x = "Difference in Median Win Percentage (Home  - Away)",
    y = "Frequency"
  )

```

 
Explanation of Plot:

The histogram shows the distribution of permuted differences in median win percentages between home and away games, simulating what we’d expect to see under the null hypothesis (no real difference between home and away performance). Each bar represents the frequency of permuted differences from our 1,000 random shuffles of the home and away labels. The red dashed line on the far right marks the observed difference in median win percentage, which is approximately 31.17%.


Summary of Findings:

The observed difference in median win percentages between home and away games is 31.17%, indicating that teams tend to perform better at home by a substantial margin. The two-sided p-value of 0 confirms that this observed difference is highly statistically significant. None of the permuted differences reached or exceeded the observed value, strongly suggesting that the observed home-court advantage is not due to random variation.

This analysis provides strong evidence that playing at home has a positive and significant impact on NCAA teams' win percentages. Factors such as crowd support, familiarity with the environment, and the absence of travel fatigue may contribute to this advantage. The findings show that home-court advantage is likely a real and influential factor in game outcomes.